Exploring Uncharted Territories - Predicting Activty Cliffs in Structure-Activity Landscapes

### 2012-07-27 paper.tex
some minor typos

Note that the predictions on the low end of the SALI values are not as important as those at the
high end - simply because low SALI values correspond to small activity cliffs, which are likely not
very interesting. Given that observation, it is encouraging to observe that all three models perform
relatively well at the higher end of the SALI spectrum. For this example, the three most significant
activity cliffs are in fact not very significant activity cliffs in an absolute sense - the $T_c$
for the three pairs are just 0.2, 0.30 and 0.29, though the activity differences were 4.93, 5.0 and
5.11 log units respectively.  But more importantly, it is clear from all three panels that the there
appears to be two different systematic prediction errors - smaller SALI values are overestimated
whereas larger values are underestimated.  A possible explanation for this behavior is that the
descriptors do not sufficiently differentiate observations with low and high SALI values. In other
words, the descriptor profile for observations with low SALI values is sufficiently similar to those
with high SALI values. To test this out, we split the Cavalli dataset into two groups - low and high
- using the median SALI value as the cutoff. We then evaluated the pairwise descriptor distance
distributions for each each of the groups, using the Euclidean distance between all descriptors in
the pool. Figure \ref{fig:cavalli-desc-dist} indicates that the distance distributions are
essentially identical. Though random forest models employ a subset of the descriptor pool for each
tree, the same reasoning can be applied to these subsets. These trends appear in the other datasets
(Figure \ref{fig:rfmodel-chembl}) but to varying degrees and not always consistently. For example,
in the Costanzo dataset, we see that the higher SALI values are somewhat underestimated. The concept
of descriptor similarity could also be applied to these cases. But a more fundamental explanation of
this behavior is not available at this time.

@@

\ctable[cap={}, caption={A comparison of the pairwise descriptor distribution for low and high SALI
  values in the Cavalli dataset. This figure employed the descriptor values generated using the
  $f_{diff}$ aggregation function and Euclidean distances were evaluataed using all the descriptors
  in the pool. The \emph{low} group is defined as those observations with SALI
  less than 2.03 and the remainder is assigned to the \emph{high} group.},
label={fig:cavalli-desc-dist}, botcap, figure]
{c}
{}
{
\includegraphics[width=\linewidth]{cavalli-desc-dist}
}

>>>>

Note that the predictions on the low end of the SALI values are not as important as those at the
high end - simply because low SALI values correspond to small activity cliffs, which are likely not
very interesting. Given that observation, it is encouraging to observe that all three models perform
relatively well at the higher end of the SALI spectrum. For this example, the three most significant
activity cliffs are in fact not very significant activity cliffs in an absolute sense - the $T_c$
for the three pairs are just 0.2, 0.30 and 0.29, though the activity differences were 4.93, 5.0 and
5.11 log units respectively.  But more importantly, it is clear from all three panels that the there
appears to be two different systematic prediction errors - smaller SALI values are overestimated
whereas larger values are underestimated.  A possible explanation for this behavior is that the
descriptors do not sufficiently differentiate observations with low and high SALI values. In other
words, the descriptor profile for observations with low SALI values is sufficiently similar to those
with high SALI values. To test this, we split the Cavalli dataset into two groups - low and high
- using the median SALI value as the cutoff. We then evaluated the pairwise descriptor distance
distributions for each each of the groups, using the Euclidean distance between all descriptors in
the pool. Figure \ref{fig:cavalli-desc-dist} indicates that the distance distributions are
essentially identical. Though random forest models employ a subset of the descriptor pool for each
tree, the same reasoning can be applied to these subsets. These trends appear in the other datasets
(Figure \ref{fig:rfmodel-chembl}) but to varying degrees and not always consistently. For example,
in the Costanzo dataset, we see that the higher SALI values are somewhat underestimated. The concept
of descriptor similarity could also be applied to these cases. But a more fundamental explanation of
this behavior is not available at this time.

@@

\ctable[cap={}, caption={A comparison of the pairwise descriptor distribution for low and high SALI
  values in the Cavalli dataset. This figure employed the descriptor values generated using the
  $f_{\textrm{diff}}$ aggregation function and Euclidean distances were evaluated using all the descriptors
  in the pool. The \emph{low} group is defined as those observations with SALI
  less than 2.03 and the remainder are assigned to the \emph{high} group.},
label={fig:cavalli-desc-dist}, botcap, figure]
{c}
{}
{
\includegraphics[width=\linewidth]{cavalli-desc-dist}
}
